AusTraits -- a curated plant trait database for the Australian flora

### 2020-11-03 ms.Rmd
spelling, grammar, usage

By providing a harmonised and curated dataset on `r austraits$traits$trait_name %>% n_distinct()` plant traits, AusTraits contributes substantially to filling the gap in Australian and global biodiversity resources. Prior to the development of AusTraits, data on Australian plant traits existed largely as a series of disconnected datasets collected by individual laboratories or initiatives \citep[but see][]{guerin2017opportunities}. We envision AusTraits as an on-going collaborative initiative for easily archiving and sharing trait data to accelerate our collective understanding the Australian flora. Open access to a comprehensive resource like this will generate significant new knowledge about the Australian flora across multiple scales of interest, as well as reduce duplication of effort in the compilation of plant trait data, particularly for research students and government agencies seeking to access information on traits.

@@

The schema of AusTraits broadly follows the principles of the established Observation and Measurement Ontology \citep{madin2007ontology} in that, where available, trait data are connected to contextual information about the collection site (e.g. location co-ordinates, soil conditions, number of replicates) and information about the methods used to derive measurements (e.g. number of replicates, equipment used where available). The finished database contains 11 elements, as described in Table 1. This format was developed to include information about the trait measurements, taxa sampled, the methods used, sites, contextual information, the peopled involved, and citation sources.

@@

Static versions of the AusTraits, including version`r austraits$build_info$version` used in this descriptor, are available on the project website (https://github.com/traitecoevo/austraits) and Zenodo (Data Citation 1). The latest data can also be downloaded directly from the project website. As validation (see Technical Validation, below) and data entry is ongoing, users are recommended to pull data from the static releases, to ensure results in their downstream analyses remain consistent as the database is updated.

@@

Comparing coverage in AusTraits to the global database TRY, there were `r aus_try %>% filter(n_austraits >0, n_try >0)%>% nrow()` traits overlapping. Of these, AusTraits tended to contain records for more taxa, but not always (Figure 5). Multiple traits had more than 10 times the number of taxa represented in AusTraits. However, there were more records in TRY for `r aus_try %>% filter(n_try > n_austraits, n_austraits >0) %>% nrow()` traits, in particular quite physiological leaf traits. Many traits were not overlapping between the two databases (Figure 5). We noted that AusTraits includes more seed and fruit nutrient data; possibly reflecting the interest in Australia in understanding how fruit and seeds are provisioned in nutrient depauperate environments. AusTraits includes more categorical values, especially variables documenting different components of species’ fire response strategies, reflecting the importance of fire in shaping Australian communities and the research to document different strategies species have evolved to succeed in fire-prone environments.

@@

We implemented several strategies to maintain data quality. First, we conducted detailed review of each source based on bespoke reports. A report is generated for each source, showing all data and metadata, and is checked by both an AusTraits curator and, where possible, the original contributor for accuracy. In these reports, observations for each trait are plotted against all other values for the trait in AusTraits. This allows AusTraits curators and contributors to quickly visually check the data to ensure it plots against other trait values as expected. Corrections suggested by contributors are combined back into AusTraits and made available with the next release.

@@

Each data release is a compressed folder containing separate files for each of the major components of AusTraits (Table 2). Detailed examples for analysis are available on the project website.

Matched against location records.

Phylogenetic tree constructed by pruning  \citep{Smith-2018} using package `V.PhyloMaker` \citep{Jin-2020} and visualising using the package `ggtree` \citep{Yu-2017}

@@

All code, raw and compiled data are hosted within a GitHub repository under the Trait Ecology and Evolution organisation (see https://github.com/traitecoevo/austraits.build, https://github.com/traitecoevo/austraits). Sequential versionscompiled dataset are also being archived within Zenodo \citep{falster_2019}. The archived material includes all data sources and code for rebuilding the compiled dataset.

@@

![The phylogenetic representation of traits in the AusTraits database for the subset of 2000 randomly sampled taxa. The heatmap colour intensity denotes the number of traits measured within family for each plant tissue. The most widespread family names (with more than ten taxa) are labelled on the edge of the tree.](figures/austraits_fig_austraits_phylogenetic_coverage.pdf)

@@

![The number of taxa with trait records in AusTraits and global TRY database (asccessed 28 May 2020). Each point shows a seperate trait. ](figures/austraits_fig_austraits_try_comparison.pdf)

>>>>

By providing a harmonised and curated dataset on `r austraits$traits$trait_name %>% n_distinct()` plant traits, AusTraits contributes substantially to filling the gap in Australian and global biodiversity resources. Prior to the development of AusTraits, data on Australian plant traits existed largely as a series of disconnected datasets collected by individual laboratories or initiatives \citep[but see][]{guerin2017opportunities}. We envision AusTraits as an on-going collaborative initiative for easily archiving and sharing trait data to accelerate our collective understanding of the Australian flora. Open access to a comprehensive resource like this will generate significant new knowledge about the Australian flora across multiple scales of interest, as well as reduce duplication of effort in the compilation of plant trait data, particularly for research students and government agencies seeking to access information on traits.

@@

The schema of AusTraits broadly follows the principles of the established Observation and Measurement Ontology \citep{madin2007ontology} in that, where available, trait data are connected to contextual information about the collection site (e.g. location coordinates, soil conditions, number of replicates) and information about the methods used to derive measurements (e.g. number of replicates, equipment used where available). The finished database contains 11 elements, as described in Table 1. This format was developed to include information about the trait measurements, taxa sampled, the methods used, sites, contextual information, the people involved, and citation sources.

@@

### Access

As an evolving data product, successive versions of AusTraits are being released, containing updates and corrections. Versions are labeled using semantic versioning to indicate the change between versions \citep{Falster-2019}. Static versions of the AusTraits, including version`r austraits$build_info$version` used in this descriptor, are available on the project website (https://github.com/traitecoevo/austraits) and Zenodo (Data Citation 1). The latest data can also be downloaded directly from the project website. As validation (see Technical Validation, below) and data entry is ongoing, users are recommended to pull data from the static releases, to ensure results in their downstream analyses remain consistent as the database is updated.

@@

Comparing coverage in AusTraits to the global database TRY, there were `r aus_try_comparison %>% filter(n_austraits >0, n_try >0)%>% nrow()` traits overlapping. Of these, AusTraits tended to contain records for more taxa, but not always (Figure 5). Multiple traits had more than 10 times the number of taxa represented in AusTraits. However, there were more records in TRY for `r aus_try_comparison %>% filter(n_try > n_austraits, n_austraits >0) %>% nrow()` traits, in particular quite physiological leaf traits. Many traits were not overlapping between the two databases (Figure 5). We noted that AusTraits includes more seed and fruit nutrient data; possibly reflecting the interest in Australia in understanding how fruit and seeds are provisioned in nutrient-depauperate environments. AusTraits includes more categorical values, especially variables documenting different components of species’ fire response strategies, reflecting the importance of fire in shaping Australian communities and the research to document different strategies species have evolved to succeed in fire-prone environments.

@@

We implemented three strategies to maintain data quality. First, we conducted a detailed review of each source based on bespoke report, showing all data and metadata, by both an AusTraits curator and the original contributor (where possible). Observations for each trait are plotted against all other values for the trait in AusTraits, allowing quick identification of outliers. Corrections suggested by contributors are combined back into AusTraits and made available with the next release.

@@

Each data release is available in multiple formats. First, as a compressed folder containing text files for each of the main components. Second, as a compressed R object, enabling easy loading into R for those using that platform.

Data is released under a CC-BY license enabling reuse with attribution -- being citation of this descriptor and, where possible, original sources.

Using the taxon names aligned with the APC, data can be eaily queried against location data from the Atlas of Living Australia. To create the phylogenetic tree in Figure 5, we pruned a master tree for all higher plants \citep{Smith-2018} using the package `V.PhyloMaker` \citep{Jin-2020} and visualising via `ggtree` \citep{Yu-2017}.

@@

All code, raw and compiled data are hosted within GitHub repositories under the Trait Ecology and Evolution organisation (see https://github.com/traitecoevo/austraits.build, https://github.com/traitecoevo/austraits). The archived material includes all data sources and code for rebuilding the compiled dataset. The code used to produce this paper is available at http://github.com/traitecoevo/austraits_ms.

@@

![The phylogenetic representation of traits in the AusTraits database for the subset of 2000 randomly sampled taxa. The heatmap colour intensity denotes the number of traits measured within a family for each plant tissue. The most widespread family names (with more than ten taxa) are labelled on the edge of the tree.](figures/austraits_fig_austraits_phylogenetic_coverage.pdf)

@@

![The number of taxa with trait records in AusTraits and global TRY database (accessed 28 May 2020). Each point shows a separate trait. ](figures/austraits_fig_austraits_try_comparison.pdf)
